---
title: "Hands-On: Fitting a Graded Response Model"
subtitle: "with NaDiRA dataset"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: none
    df_print: paged
---

## Welcome!

This is the hands-on companion to *"Introduction to IRT's Graded Response Model"*. It mirrors the workflow from [the tutorial paper](https://onlinelibrary.wiley.com/doi/abs/10.1002/ijop.13265) this workshop is based on, but on a different dataset.

> **Two ways to run this document**
>
> This document can run on the **snippet of real NaDiRA dataset** (for the workshop room) or on **simulated data calibrated to resemble it**.
>
> - **In the workshop:** setting `USE_SIMULATED_DATA <- FALSE` — will read the real dataset.
>   + Before running `.qmd` or `.Rmd` document with the real dataset, please create a new folder `data/` in your project root, then keep the dataset there.
> - **If you don't have any access to the dataset:** `USE_SIMULATED_DATA <- TRUE`, which generates a synthetic dataset from item parameters estimated on the real data, so the shape and difficulty of the exercise carry over, but every response is simulated with `mirt` package. There is no real participants' data in it.

We'll go through 10 steps together: meet the dataset, check dimensionality, fit the GRM, read its output, inspect residual correlations, and think about what it means for your own scales.

> **If you're running the simulated dataset**
>
> The numbers and conclusions quoted in Steps 8 and 9 describe the **real NaDiRA data** and won't replicate under `USE_SIMULATED_DATA <- TRUE`. `simdata()` generates responses that exactly satisfy the fitted GRM, so the simulated model and item fit (Step 8) comes out essentially perfect (no misfitting model or items at all), and it has no local dependence built in at all (the problematic items highlighted in Step 9 won't appear). Steps 2–7 will look similar, though not identical, either way.

## Step 0: Setup — packages

If you already have these installed, skip this chunk.

```{r, eval=FALSE}
install.packages(
    c("tidyverse", # data wrangling
      "psych", # descriptive stats & unidimensionality checks
      "pak", # to install a GitHub-only package (ggmirt)
      "mirt", # the main GRM engine
      "skimr", # quick look at data distributions
      "haven" # to import the NaDiRA STATA .dta file
    ),
    dependencies = TRUE
)

pak::pak("masurp/ggmirt") # ggmirt available only on github
```

Now activate them.

```{r, message=FALSE}
library(tidyverse)
library(psych)
library(mirt)
library(ggmirt)
library(skimr)
library(haven)
```

## Step 1: Meet the dataset

### The swap-in code chunk

Everything dataset-specific lives here. This is the only chunk you should need to touch to move between the real data and the simulated version.

```{r}
USE_SIMULATED_DATA <- FALSE # set TRUE if you don't have an access to the dataset

DATA_PATH <- "data/NaDiRa_Teaching_w0_clean.dta" # path to the dataset don't forget to create a new folder `data/` in your project root, then keep the dataset file there

ITEM_COLS <- c(
    "ker_diskerf_service", "ker_diskerf_respekt", "ker_diskerf_nernst",
    "ker_diskerf_angst", "ker_diskerf_bedbel", "ker_diskerf_beleid",
    "ker_diskerf_angriff"
)

N_CATEGORIES <- 6 # response categories per item (1-6)
```

### The items

This is question 3 of the NaDiRA survey instrument: *"Kommen wir nun zu Ihren Erfahrungen im täglichen Leben. Wie häufig passieren Ihnen in Ihrem Alltag die folgenden Dinge?"* ("Now let's turn to your everyday experiences. How often do the following things happen to you in your daily life?")

| Variable name | 🇩🇪 German original | 🇬🇧 English |
|------------------------|------------------------|------------------------|
| `ker_diskerf_service` | Sie erhalten schlechteren Service als andere. | You receive poorer service than other people. |
| `ker_diskerf_respekt` | Sie werden mit weniger Respekt behandelt als andere. | You are treated with less respect than other people. |
| `ker_diskerf_nernst` | Jemand verhält sich so, als würde sie/er Sie nicht ernst nehmen. | Someone acts as if they don't take you seriously. |
| `ker_diskerf_angst` | Jemand verhält sich so, als habe sie/er Angst vor Ihnen. | Someone acts as if they're afraid of you. |
| `ker_diskerf_bedbel` | Sie werden bedroht/belästigt. | You are threatened/harassed. |
| `ker_diskerf_beleid` | Sie werden beleidigt/beschimpft. | You are insulted/being called names. |
| `ker_diskerf_angriff` | Sie werden körperlich angegriffen. | You are physically attacked. |

The response options in the questionnaire run left to right: **Daily/*Täglich* (1) · At least once per week/*Mindestens einmal in der Woche* (2) · At least once per month/*Mindestens einmal im Monat* (3) · Several times a year/*Mehrmals im Jahr* (4) · Less often/*Seltener* (5) · Never/*Nie* (6)**.

> **No need to reverse the scores**
>
> You are given a cleaned dataset in which all the scale data has been reversed. So, the larger scores now reflect more intense experience of discrimination, and *vice versa*.

### Importing and cleaning the dataset

Run this code chunk to simulate/generate a synthetic dataset. If you have access to a real dataset, you can also run this code chunk to clean and prepare it for analysis.

```{r}
if (USE_SIMULATED_DATA) {
    # ---- calibrated simulation: parameters estimated once on the real data ----
    # (a = discrimination, d = intercepts; this array reproduces the item
    # difficulty/discrimination pattern of the real dataset without using any
    # real participants' answers)
    set.seed(2026)

    a_par <- c(2.51, 3.48, 2.87, 1.21, 1.68, 1.97, 1.13) # discrimination params
    d_par <- matrix(                                     # intercepts (d = -a * b)
         c(
            # d1        d2        d3        d4        d5
            1.9284,  -1.5920,  -3.0283,  -5.0120,  -6.7521,  # service
            2.5426,  -1.7032,  -3.6920,  -5.5658,  -7.7253,  # respekt
            3.0612,  -0.7804,  -2.6710,  -4.2892,  -6.9545,  # nernst
            -0.3967,  -2.3890,  -3.2141,  -4.2315,  -5.4047,  # angst
            -0.5662,  -2.9232,  -4.3999,  -5.5944,  -6.6881,  # bedbel
            0.3987,  -2.2918,  -3.8572,  -5.1107,  -7.8703,  # beleid
            -2.0059,  -4.3481,  -5.0828,  -5.4529,  -6.7298   # angriff
        ),
        nrow = length(ITEM_COLS), ncol = N_CATEGORIES - 1, byrow = TRUE
    )

    data <- simdata(a = a_par, d = d_par, N = 970, itemtype = "graded") + 1
    data <- as.data.frame(data)
    colnames(raw) <- ITEM_COLS
} else {
    data <- read_dta(DATA_PATH) %>%
        select(all_of(ITEM_COLS)) %>%
        mutate(across(everything(), as.numeric)) # drop haven_labelled class, keep plain numeric codes
}
```

We already have a cleaned dataset (*N* = `r nrow(data)`).

## Step 2: Quick data check

```{r}
describe(data)
skim(data)
```

Look at the `hist` column — frequency-of-negative-experience items like these are typically right-skewed (most people report "Nie"/"Seltener"). `angriff` (physical attack) should look the most extreme; the everyday-slight items (`nernst`, `respekt`, `service`) should look comparatively less skewed.

That skew is exactly why we'll use a **polychoric correlation matrix** (instead of Pearson's or Spearman) for the dimensionality check below.

And that's exactly what we see: `angriff` is by far the most skewed item (skew = 4.21, mean = 1.20, so almost everyone answers "Nie"), while `nernst` has the lowest skew (skew = 0.79) — its mean (2.56) is actually the highest of the seven, meaning this everyday slight is common enough that responses spread more evenly across the scale instead of piling up at one end.

The severe/rare items and the everyday/common items sit at opposite ends, which is a good sign the item set spans a real range of severity rather than all measuring the same narrow slice of experience.

## Step 3: Dimensionality

We already ran this exact logic in the factor analysis session. We will do a fast check to confirm a single $\theta$ underlies this item pool.

With only 6 response categories, Pearson's/Spearman correlations understate how strongly these items actually relate to each other, so a **polychoric** correlation matrix is the more appropriate input here for factor analysis ([Flora & Curran, 2004](https://psycnet.apa.org/record/2004-21445-005); recommended specifically for 4–7 category Likert data by [Guo & Choi, 2023](https://journals.sagepub.com/doi/10.1177/00131644221111838)). `psych::irt.fa()` computes it for us automatically:

```{r}
efa <- irt.fa(data, nfactors = 1, fm = "minres", plot = FALSE)
print(efa$fa)
```

Here, it shows that Proportion Var = .53 in the EFA analysis above, with loadings between .57 and .85. Every item is solidly on the one factor, none of them weak or borderline.

We can also run a parallel analysis if we are unsure about the number of $\theta$ underlies our item pool ([Liu et al., 2007](https://onlinelibrary.wiley.com/doi/10.1002/j.2333-8504.2007.tb02083.x)) by running this script:

```{r}
set.seed(2026) # fixes the simulated reference line so the plot is identical on every render
a <- fa.parallel(polychoric(data)$rho, n.obs = nrow(data), fm = "minres", main = "Scree Plot Everyday Discrimination Scale - NaDiRA Test Dataset")
a$fa.values
```

If the proportion of variance explained by the first factor is comfortably above 50% and the scree plot elbows sharply after one factor, unidimensionality is reasonably supported, so we can proceed to the GRM.

The scree plot backs up the EFA results: eigenvalues 3.70 then a sharp drop to 0.5. Note that `fa.parallel()`'s own text output will say it "suggests the number of factors = 2" — the second eigenvalue does clear the simulated reference line, driven mainly by the three overt-hostility items (`bedbel`, `beleid`, `angriff`) pulling slightly apart from the rest.

The two factors are still substantially correlated (*r* ≈ .57), and the first factor alone accounts for the bulk of the variance, so a single-factor GRM remains a reasonable working approximation. Keep this item grouping in mind, though, since it resurfaces in Step 9.

## Step 4: Fit the GRM

```{r}
model <- paste0("theta = 1-", length(ITEM_COLS)) # specifying the model
model # we can see here the model contains one LV (theta) with 7 items
fit <- mirt(data = data, model = model, itemtype = "graded", SE = TRUE, verbose = FALSE) # now fit the model
coefs <- coef(fit, IRTpars = TRUE, simplify = TRUE) # keep item parameters in a list
print(coefs) # print item parameters
```

`a` is discrimination; `b1`–`b5` are the five thresholds (six categories → five boundaries). Higher `a` = an item that differentiates people more sharply; the spread of `b` values tells us *where* on $\theta$ each item is doing its work.

In this scale, `respekt` ("less respect") has the highest discrimination (*a* = 3.48), which makes it the sharpest differentiator of the seven, while `angriff` ("being physically attacked") has the lowest (*a* = 1.12).

`angriff` ("physically attacked") also stands out on its thresholds: they run from 1.78 up to 5.98, far higher than any other item, meaning it takes an extremely high $\theta$ before "Täglich" becomes a plausible response, which tracks with how rare that experience actually is.

## Step 5: Item probability curves

```{r}
tracePlot(fit, title = "Item Probability Functions") + labs(color = "Response Category")
```

Thresholds are where the probability of scoring in category k or higher reaches 50%. Curves that overlap a lot, or peak only at one end of $\theta$, are telling you that item is only informative for part of the trait range.

Pick out `angriff` in the grid: because its thresholds sit so far to the right, the "Nie" curve dominates almost the entire visible range. The other five categories barely get a look-in until $\theta$ is very high. Compare that to `nernst` and `respekt`, whose curves are bunched much closer to the middle, it's doing useful work across a wider, more central stretch of $\theta$.

## Step 6: Item and test information

This is the information that traditional factor analysis can't give: *where* exactly our scale is precise.

```{r}
itemInfoPlot(fit, facet = TRUE, title = "Item Information Functions")
```

> **Why some curves have multiple peaks (i.e., multimodal)?**
>
> A yes/no item only has one "tipping point" i.e., the trait level where people switch from one answer to the other, so it's only ever most useful for one type of participants. These items have six response categories, so each one has several tipping points or "steps" as in [the staircase illustration](intro.html#/the-staircase-illustration) (e.g., "rarely" to "sometimes", "sometimes" to "often", and so on). If those tipping points are spread far apart along the trait, the item ends up being especially useful for more than one type of participant, therefore showing up as more than one peak instead of just one.

```{r}
testInfoPlot(fit, title = "Test Information Function")
```

Compare the shape of the test information curve to where we'd expect our target population to actually sit on $\theta$. A curve that peaks somewhere our participants rarely fall is a scale that's precise in the wrong place.

Here, test information peaks around $\theta \approx 1.25$ (info $\approx 10.7$) and stays moderately high out to $\theta \approx 3$ (info $\approx 6.0$), but falls off sharply below $\theta \approx -1$ (info $\approx 1.2$ at $\theta = -2$, about a tenth of the peak).

## Step 7: Reliability across $\theta$

**Marginal reliability** is the IRT analogue of classical test-theory reliability: it integrates the test information function over an *assumed* $\theta$ distribution (normal, by default) to get a single reliability estimate for the population as a whole.

```{r}
marginal_rxx(fit)
```

**Empirical reliability** instead uses the actual estimated $\theta$ scores and their standard errors for *this* sample, so it reflects the observed spread of scores. It's the ratio of true-score variance to observed-score variance, computed directly from the EAP estimates (or factor scores, in CTT's term) and their SEs.

```{r}
theta_se <- fscores(fit, full.scores.SE = TRUE)
empirical_rxx(theta_se)
```

```{r}
# using mirt's own plot() rather than ggmirt::conRelPlot() here -- for some reason,
# mirt/ggmirt version combination conRelPlot() crashes R outright
# mirt's native rxx plot shows the same thing
plot(fit, type = "rxx", main = "Reliability Across the θ Continuum")
```

Marginal and empirical reliability should be close. The plot shows us *where* reliability holds up (commonly interpreted as adequate above ~.80) and where it doesn't, and usually the tails.

Here they are: marginal rxx = .870, empirical rxx = .873, which is what we want. Reliability stays $\geq$ .80 across roughly $\theta$ = -1.29 to 3.18, lining up with the information plot above.

> **What does this mean?**
>
> In practice: this scale is precise at telling apart people with moderate-to-high discrimination-experience levels, but has little to say about fine-grained differences among people reporting very few or no experiences, which points out exactly the "blind spot" idea from the slides.

## Step 8: Model and item fit parameters

Now, let us check the model and item fit parameters.

```{r}
M2(fit, type = "C2") # we use C2 because we're dealing with polytomous data
```

Here the picture is mixed: RMSEA = .118 (90% CI .104-.133) is well above the ≤.05 "good fit" guideline, but CFI = .951 meets the "good fit" guideline, SRMSR = .078 is rather acceptable, and TLI = .926 is borderline.

> SRMSR here is the reliable signal of *how much* misfit there practically is, and if it's less than 0.08, then it may indicate that some items are misfitting, but perhaps not so severe.

Now, let us see the item fit statistics.

```{r}
itemfit(fit)
```

Rough guide: RMSEA ≤ .05 and non-significant item-level tests are what we want; **but** avoid over-interpret significance with very large samples, since almost any model will show *some* misfit anyway. RMSEA tends to be an unreliable indicator with only a handful of items and few degrees of freedom (df = 14 here). The most important question would be: how severe is the misfit and can we reasonably tolerate it?

> **Item misfit**
>
> `itemfit()`'s default test ($S\text{-}X^2$) checks $H_0$: *this item's responses fit the model's predictions*. A **significant *p*-value flags misfit** (we reject $H_0$); a non-significant one is what we want: this means the observed responses don't deviate from what the GRM expects for that item.

At the item level, four of seven items (`nernst`, `angst`, `bedbel`, `beleid`) show significant misfit (*p* < .05); with *N* = `r nrow(data)` that's not simply a huge-sample artifact the way it would be with, say, 9,000+ respondents, so it's worth a closer look.

## Step 9: Inspecting residual correlations (local dependence)

> If global (model) fit tells us *whether* there's a problem; residual correlations tell us *where* the problems likely lie.

If two items still correlate once $\theta$ is accounted for, they're carrying some shared content the model isn't capturing, which is also called **local dependence (LD)**. We'll use Yen's $Q_3$ statistic, which is recommended over the LD $\chi^2$ statistic specifically for polytomous data like this.

To flag problematic items, we may apply |$Q_3$| > 0.2 critical value, that is commonly used ([Chen & Thissen, 1997](https://journals.sagepub.com/doi/abs/10.3102/10769986022003265)), as a first-pass screen, but see the negative-bias caveat below before treating it as a hard cutoff.

> |$Q_3$| > 0.2 is a commonly used rule of thumb in practice, but as with many other rules of thumb, some people question the accuracy (e.g., [Nason & DeMars, 2025](https://onlinelibrary.wiley.com/doi/abs/10.1111/jedm.12432))

```{r}
q3 <- residuals(fit, type = "Q3", verbose = FALSE) # asking mirt to compute q3
round(q3, 2) # round the value to only two decimals
```

A couple of things worth knowing before reading this table at face value.

First, $Q_3$ has a well-documented *negative* bias, especially with few items: because $\theta$ is estimated from the same items whose residuals are being correlated, the "no local dependence" baseline isn't *exactly* 0, but it's somewhat below zero ([Yen, 1984, p. 127](https://journals.sagepub.com/doi/abs/10.1177/014662168400800201); Marais, 2013, p. 121, in [Christensen, Kreiner & Mesbah (Eds.), 2013](https://onlinelibrary.wiley.com/doi/book/10.1002/9781118574454)). Here that mean off-diagonal $Q_3$ is -.113.

Most of the negative values in this residual correlation matrix are the expected statistical byproduct of a handful of strongly-loading items sharing one factor. Instead of comparing the $Q_3$ values to 0, we should evaluate them relative to the mean-off $\bar{Q}_3$ = -.113 (centered as $Q_{3d} = Q_3 - \bar{Q}_3$).

```{r}
baseline <- mean(q3[upper.tri(q3)]) # calculating mean off-diagonal, to see the "no-LD" baseline
as.data.frame(q3) %>%
  rownames_to_column("item1") %>%
  pivot_longer(-item1, names_to = "item2", values_to = "Q3") %>%
  filter(item1 < item2) %>%
  mutate(Q3d = Q3 - baseline) %>%
  filter(Q3 > 0, Q3d > 0.2) # flag problematic items that have more paired correlations more than 0.2 from the baseline (-0.113)
```

All three pairs above exceed the common rule of thumb threshold of $Q_{3d} > +0.20$ above the baseline. This may indicate that experiences of overt verbal hostility (`beleid`), threats (`bedbel`), and physical assault (`angriff`) share "nuisance variance" (i.e., "unknown source") beyond general everyday discrimination, and this is likely because they frequently co-occur during the same hostile encounters.

However, in my view, none of this is severe enough to justify dropping an item.

If the same pattern shows up on the full NaDiRA sample, alternative reasonable next steps are:

* Modeling the "overt hostility" items (`bedbel`, `beleid`, `angriff`) as a specific testlet or secondary factor alongside the general discrimination factor (i.e., a bifactor structure).
* Refining item wording in future survey waves to distinguish distinct situational contexts, or
* **as a last resort**, removing items.
    + Removing an item only if testlet approaches fail and parameter distortion remains an issue, keeping in mind the potential loss in content coverage.
    + Removing items should stay a last resort, because while it can improve model fit it also narrows the scope of what the scale actually measures.

## Try it yourself

A few things to explore independently after this session:

- Drop the item with the lowest discrimination (`a`) from `model` and re-fit. How much does the test information curve change?
- If you were designing a scale for measuring people's everyday discrimination experiences for a large-scale survey like NaDiRA, which part of $\theta$ would you most want it to be reliable in, and does anything here suggest you'd need more items there?
- If you had the full NaDiRA sample instead of this teaching extract, which item from Step 9 would you investigate first and would you reword, parcel, or drop?

## Contact

Questions or feedback: [amelia.zein@cais-research.de](mailto:amelia.zein@cais-research.de) or [amelia.zein@psikologi.unair.ac.id](mailto:amelia.zein@psikologi.unair.ac.id)

## Session Info

```{r}
sessionInfo()
```
